---
pagetitle: Binomial Trend Detection - Identifying Ongoing Shifts in Data
description-meta: Learn how the binomial trend detection method can identify ongoing trends in data over time by comparing week-over-week changes. This approach provides a statistical basis for detecting trends and overcoming the limitations of traditional rolling averages.
toc: false
lightbox: true
execute: 
  freeze: auto
---

# Binomial Trend Detection

Detecting trends and anomalies can be challenging, but rolling average benchmarking is a common method that smooths out fluctuations by comparing each new data point to the average of a specified number of previous values.

While a rolling average can help highlight a metric trend, it can have limitations:

- Sensitivity to window size
- Lag in response
- Difficulty accounting for varying granularities of seasonality or sudden changes

An alternative method that can help identify ongoing shifts in trends over time is the **binomial trend detection approach**, which uses week-over-week (WoW) comparisons to detect changes.

The binomial method analyzes week-over-week changes in data using a 14-day rolling window. It counts the number of days with positive or negative changes within this window and applies the binomial distribution to assess the probability of a consistent trend. A predefined threshold, such as 12 successes out of 14 trials (days) in the same direction, is used to identify statistically significant trends.

| **Step** | **Details** |
|----|----|
| **Calculate WoW Deltas** | Compute week-over-week changes (deltas) for the metric of interest, determining whether the metric has increased or decreased each day. |
| **Rolling 2-Week Window** | Analyze data within a 14-day rolling window, smoothing out short-term fluctuations while responding to recent changes. |
| **Counting Positive and Negative Days** | Count the number of days with positive or negative WoW deltas within the 14-day window to assess recent performance trends. |
| **Binomial Distribution** | Treat each day as an independent trial with a 50% chance of being positive or negative, using the binomial distribution for statistical analysis. |
| **Trend Threshold** | Set a threshold for statistical significance (e.g., 12 or more positive/negative days out of 14) to ensure the trend is consistent and not random. |
: Method Steps {.striped}

```{r, binom-test}
#| code-overflow: wrap
binom.test(12, 14, 1/2, alternative = "greater")
```


```{r, binom-code}
#| warning: false
#| code-overflow: wrap
#| code-fold: true
library(ggplot2)
library(showtext)
library(ggtext)

showtext_auto()
showtext_opts(dpi = 300)

font_add_google(name = "Roboto", family = "Roboto")
font_1 <- "Roboto"

n <- 14 
p <- 1/2  
x_obs <- 12  

binom_test <- binom.test(x_obs, n, p, alternative = "greater")

p_value <- binom_test$p.value
conf_int <- binom_test$conf.int

p_ge_x_obs <- sum(dbinom(x_obs:n, size = n, prob = p))

data <- data.frame(
  x = 0:n,
  probability = dbinom(0:n, size = n, prob = p),
  color = ifelse(0:n >= x_obs, "#C0392B", "#30394F")  
)

binom_visual <- ggplot(data, aes(x = x, y = probability, fill = color)) +
  geom_bar(stat = "identity") +
  scale_fill_identity() +  
  scale_x_continuous(breaks = 0:n) +  
  labs(
    title = "Binomial Distribution (n = 14, p = 0.5)",
    x = "Number of Successes",
    y = "Probability"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(10, 10, 10, 10, "mm"),
    axis.text = element_text(family = font_1, size = 7),
    axis.title = element_text(family = font_1, size = 7),
    axis.title.x = element_text(margin = margin(5, 0, 0, 0, 'mm')),
    axis.title.y = element_text(margin = margin(0, 5, 0, 0, 'mm')),
    plot.title = element_text(family = font_1, size = 10)
  ) +
  annotate(
    geom = 'richtext',
    x = n+1,
    y = max(data$probability) * 0.9,
    label = paste0(
      "<span style='color:#C0392B; font-size:8pt;font-family:Roboto;'>",
      "Threshold ≥ 12 trials (days)", "<br>",
      "P(X ≥ ", x_obs, ") = ", round(p_ge_x_obs, 4), "<br>",
      "95% CI = [", round(conf_int[1], 4), ", ", round(conf_int[2], 4), "]</span>"),
    hjust = 1, fill = NA, label.color = NA
  )
```

```{r, binom-visual}
#| echo: false
#| fig-column: page
#| lightbox:
#|   group: plot
binom_visual
```

This binomial approach improves upon rolling averages by offering better responsiveness to recent trends and reducing sensitivity to day-of-week variations, leading to more reliable trend detection.
