---
pagetitle: Binomial Trend Detection
description: |
  Explore a robust method for trend detection using binomial analysis, focusing on week-over-week comparisons to identify significant patterns in data.
lightbox: true
execute: 
  freeze: auto
---

::: column-page
# Binomial Trend Detection
:::

::::: column-screen
:::: portfolio-banner
::: column-page
One common approach to highlighting trends and detecting anomalies is **rolling average benchmarking**. This method involves comparing each new data point to the average of a specified number of preceding values. By smoothing out short-term fluctuations, it becomes easier to identify deviations that may indicate a potential trend shift or anomaly.

<br>

However, this method has some limitations. It is sensitive to the chosen window size, can introduce lag, and may struggle to account for seasonality or sudden changes in the data.
:::
::::
:::::

::: {.column-page .binom}
An alternative method to detect trends more robustly is a **binomial trend detection approach**, which uses a week-over-week (WoW) comparison. This is how the method can be implemented:

| **Step** | **Details** |
|----|----|
| **Calculate WoW Deltas** | Compute the week-over-week changes (deltas) for the metric of interest. This involves determining whether the metric has increased or decreased each day. |
| **Rolling 2-Week Window** | For each day, consider the previous 14-day period (2 weeks) as a rolling timeframe for analysis. This helps smooth out short-term variations while being responsive to recent changes. |
| **Counting Positive and Negative Days** | Within the 14-day window, count the number of days where the WoW delta is positive and the number of days where it is negative. This provides insight into the recent performance pattern. |
| **Binomial Distribution** | Assume that each day has an equal 50% chance of having either a positive or negative WoW delta. Under this assumption, the 14-day window is treated as 14 independent trials, each with a 50% probability of being positive or negative. |
| **Trend Threshold** | Set a threshold to identify significant trends: for example, if 12 or more out of the 14 days show a consistent direction (either positive or negative), it is deemed statistically significant according to the binomial distribution. |


```{r, binom-test}
binom.test(12, 14, 1/2, alternative = "greater")
```

This threshold helps ensure that the observed trend is not due to random fluctuations, but rather indicates a consistent pattern.



```{r, binom-visual}
#| warning: false
#| code-fold: true
library(ggplot2)
library(showtext)
library(ggtext)

showtext_auto()
showtext_opts(dpi = 300)

font_add_google(name = "Roboto", family = "Roboto")
font_1 <- "Roboto"

n <- 14 
p <- 1/2  
x_obs <- 12  

binom_test <- binom.test(x_obs, n, p, alternative = "greater")

p_value <- binom_test$p.value
conf_int <- binom_test$conf.int

p_ge_x_obs <- sum(dbinom(x_obs:n, size = n, prob = p))

data <- data.frame(
  x = 0:n,
  probability = dbinom(0:n, size = n, prob = p),
  color = ifelse(0:n >= x_obs, "#C0392B", "#30394F")  
)

binom_visual <- ggplot(data, aes(x = x, y = probability, fill = color)) +
  geom_bar(stat = "identity") +
  scale_fill_identity() +  
  scale_x_continuous(breaks = 0:n) +  
  labs(
    title = "Binomial Distribution (n = 14, p = 0.5)",
    subtitle = "Success threshold ≥ 12 trials (days)",
    x = "Number of Successes",
    y = "Probability"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = margin(10, 10, 10, 10, "mm"),
    axis.text = element_text(family = font_1, size = 7),
    axis.title = element_text(family = font_1, size = 7),
    axis.title.x = element_text(margin = margin(5, 0, 0, 0, 'mm')),
    axis.title.y = element_text(margin = margin(0, 5, 0, 0, 'mm')),
    plot.title = element_text(family = font_1, size = 10),
    plot.subtitle = element_text(family = font_1, size = 8)
  ) +
  annotate(
    geom = 'richtext',
    x = n,
    y = max(data$probability) * 0.9,
    label = paste0(
      "<span style='color:#C0392B; font-size:8pt;font-family:Roboto;'>",
      "P(X ≥ ", x_obs, ") = ", round(p_ge_x_obs, 4), "<br>",
      "95% CI = [", round(conf_int[1], 4), ", ", round(conf_int[2], 4), "]</span>"),
    hjust = 1, fill = NA, label.color = NA
  )
```

:::

```{r, binom-visual-2}
#| echo: false
#| fig-column: page
#| lightbox:
#|   group: plot
binom_visual
```

::: {.column-page .binom}
This binomial approach balances longer rolling windows, reduces sensitivity to day-of-week variations, and adjusts to recent trends more effectively than simple rolling averages.
:::